---
title: "Growth and Institutions"
author: "Augusto Gonzalez Bonorino"
date: 'Sys.Date()'
---


```{r}
library(readxl)
library(tidyverse)

# source: https://www.rug.nl/ggdc/historicaldevelopment/maddison/releases/maddison-project-database-2023?lang=en
maddison_project <- read_excel(file.choose(), sheet = "Full data")
head(data)
```

```{r}
# data missing for Singapore (1940-1949)
filtered_maddison <- maddison_project %>%
  filter(country %in% c("United States", "Argentina", "Singapore")) %>%
  filter(year > 1850 & year < 2018)

head(filtered_maddison)
```


```{r}
# source: https://www.systemicpeace.org/inscrdata.html
polity_iv <- read_excel(file.choose())
head(polity_iv)
```

```{r}
filtered_polity_iv <- polity_iv %>%
  filter(country %in% c("United States", "Argentina", "Singapore")) %>%
  filter(year > 1900 & year < 2018) %>%
  select(ccode, country, year, democ, autoc, polity2, durable, xconst, parcomp, exrec, regtrans, sf)
```

```{r}
# download filtered_maddison to avoid dependency on full datasets
write_csv(filtered_maddison, "filtered_maddison.csv")
write_csv(filtered_polity_iv, "filtered_polity_iv.csv")
```

# FILTERED DATASETS - ANALYSIS

[Polity IV codebook](https://web.mit.edu/efogg/Public/PolisciLab/Data/p4manualv2006.pdf)

```{r}
# set working directory
setwd("C:/Users/Bonoc/OneDrive/PhD_applications/George Mason")
```


```{r}
maddison_data <- read_csv("filtered_maddison.csv")
polity_data <- read_csv("filtered_polity_iv.csv")
```

```{r}
# impute missing gdppc values for Singapore
library(zoo)

maddison_data <- maddison_data %>%
  mutate(gdppc = ifelse(is.na(gdppc) & country == "Singapore", 
                        na.approx(gdppc, rule=2, na.rm = FALSE), 
                        gdppc)) %>%
  select(country, year, gdppc)

```

```{r}
# merge the datasets
merged_data <- maddison_data %>%
  left_join(polity_data, by = c("country", "year"))
```


```{r}
library(strucchange)

# Bai-Perron test to identify structural breaks in Argentina
argentina <- merged_data %>% filter(country == "Argentina")
bp_arg <- breakpoints(log(gdppc) ~ 1, data = argentina, h = 0.15)
print(bp_arg)
plot(bp_arg)  

# Optimal breaks at 1923, 1946, 1965, 1997
```

Historical Alignment:

- 1923: Post-WWI economic turbulence and early protectionist policies. Dawn of military coup that took place in 1930.

- 1946: Rise of Peronism and import substitution industrialization (ISI).

- 1965: Political instability preceding the 1966 military coup.

- 1997: Prelude to the 2001 default and currency crisis.

## Regression analysis

I am missing institutional variables for Singapore before 1965, so I will analyze Argentina and Singapore separately using United States as a comparison baseline.

```{r}
# Argentina (full data)
model_arg <- lm(log(gdppc) ~ polity2 + durable, data = argentina)
summary(model_arg)
```

```{r}
# Singapore (1965–2017 only)
singapore <- merged_data %>% filter(country == "Singapore") %>% filter(year >= 1965)
model_sgp <- lm(log(gdppc) ~ polity2 + durable, data = singapore)
summary(model_sgp)
```

```{r}
usa <- merged_data %>% filter(country == "United States")
model_usa <- lm(log(gdppc) ~ polity2 + durable, data = usa)
summary(model_usa)
```


```{r}
# test if institutional effects differ across countries
model_global <- lm(log(gdppc) ~ polity2 * country + durable * country, 
                  data = merged_data)
summary(model_global)
```


```{r fig.width=10, fig.height=5}
# Custom labels for facets
facet_labels <- c(
  "gdppc" = "GDP per Capita",
  "polity2" = "Polity2 Score (-10 to 10)",
  "durable" = "Regime Durability"
)

# Convert data to long format and plot
merged_data %>%
  filter(country == "Argentina") %>%
  select(year, gdppc, polity2, durable) %>%
  pivot_longer(
    cols = -year,
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = year, y = value)) +
  geom_line(aes(color = metric), linewidth = 1) +
  facet_wrap(
    ~ metric,
    nrow = 1,
    scales = "free_y",
    labeller = labeller(metric = facet_labels)
  ) +
  scale_color_manual(
    values = c("gdppc" = "blue", "polity2" = "red", "durable" = "black")
  ) +
  labs(
    title = "Argentina: Key Metrics (1901–2017)",
    x = "Year",
    caption = "GDP: Maddison Project\nInstitutional Data: Polity IV\nPolity2: -10 (autocracy) to +10 (democracy).\nDurability: Years since last regime change (defined as a 3+ point change in polity2 score)."
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```


```{r fig.width=10, fig.height=5}
# Convert data to long format and plot
merged_data %>%
  filter(country == "Singapore") %>%
  select(year, gdppc, polity2, durable) %>%
  pivot_longer(
    cols = -year,
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = year, y = value)) +
  geom_line(aes(color = metric), linewidth = 1) +
  facet_wrap(
    ~ metric,
    nrow = 1,
    scales = "free_y",
    labeller = labeller(metric = facet_labels)
  ) +
  scale_color_manual(
    values = c("gdppc" = "blue", "polity2" = "red", "durable" = "black")
  ) +
  labs(
    title = "Singapore: Key Metrics (1901–2017)",
    x = "Year",
    caption = "GDP: Maddison Project\nInstitutional Data: Polity IV\nPolity2: -10 (autocracy) to +10 (democracy).\nDurability: Years since last regime change (defined as a 3+ point change in polity2 score)."
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```


```{r fig.width=8, fig.height=5}
# gdppc for each country
merged_data %>%
  ggplot(aes(x = year, y = log(gdppc), color = country)) +
  geom_line() +
  labs(title = "GDP per Capita by Country (1901–2017)", 
       x = "Year", y = "Log GDP per Capita",
       caption="Data: Maddison Project Database. Note: Singapore's pre-1965 GDPpc includes interpolated values (1940–1949 missing)") +
  theme_minimal()
```

```{r}
gdp_share
```


```{r}
# countries' gdp per capita as a share of USA gdp per capita over time
usa_gdppc <- merged_data %>% filter(country == "United States") %>% pull(gdppc)
argentina_gdppc <- merged_data %>% filter(country == "Argentina") %>% pull(gdppc)
singapore_gdppc <- merged_data %>% filter(country == "Singapore") %>% pull(gdppc)

gdp_share <- data.frame(
  year = merged_data$year,
  argentina = argentina_gdppc / usa_gdppc,
  singapore = singapore_gdppc / usa_gdppc
)

gdp_share %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = argentina, color="Argentina"), linetype = "dashed") +
  geom_line(aes(y = singapore, color="Singapore"), linetype = "dotted") +
  labs(title = "GDP per Capita as Share of USA (1901–2017)", x = "Year", y = "GDP per Capita Share") +
  scale_color_manual(values = c("Argentina" = "darkblue", "Singapore" = "darkgreen")) +
  theme_minimal()
  
```




